---
title: 'Dataset 4 : Sea Level Changes'
author: "Nic Bone"
date: "February 26, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
install.packages("tidyr")
install.packages("ggplot2")
install.packages("dplyr")
library(ggplot2)
install.packages("wesanderson")
library(wesanderson)
```

In this analysis, I will be calculating and looking at the expected mean sea level rise of the Polynesia/Micronesia/Oceania region (denoted "Pacific Islands") in the year 2050 comapared to the expected mean sea level rise of the East China Sea. I predict that the rate of increase in the Polynesia/Micronesia/Oceania will predict a significantly higher sea level increase (when controlling for different starting levels).


```{r}
# Reading in the dataset
strend <- read.csv("ECH and Pacific.csv")

head(strend)
```

```{r}
#Visualizing the distribution of trends 
ggplot(strend, aes(x=Region, y= Trend.Value, color = Region)) + geom_violin() + stat_summary(fun.y=mean, geom="point", color ="#F2AD00", shape=17, size=4) + scale_x_discrete(labels=c("East China Sea Region", "Pacific region")) +scale_color_manual(values=wes_palette(n=3, name="Darjeeling"))
```




```{r}
library(dplyr)
pac <- filter(strend, Region == ("P"))
pac

chin <- filter(strend, Region == ("C"))
chin
```
```{r}
#extrapolating the yearly trend to the year 2050

trend <- (32*(strend$Trend.Value) - (strend$Trend.Value))

ptrend <- (32*(pac$Trend.Value) - (pac$Trend.Value))

ctrend <- (32*(chin$Trend.Value) - (chin$Trend.Value))

paclevel2050 <- mutate(pac, New = ptrend)

chinlevel2050 <- mutate(chin, New = ctrend)

level2050 <- mutate(strend, New = trend)
```

```{r}
#The sea level change by the year 2050
level2050
```


Now we can see the 
```{r}
ggplot(level2050, aes(x=Region, y = New)) + geom_point()
```

```{r}
mean(chinlevel2050$New)
```

```{r}
mean(paclevel2050$New)
```
```{r}
t.test(paclevel2050$New, chinlevel2050$New)
```

The results of the t-test show that the means of each region are not sig. different from each other. 


# What it means

Although the means of the East China Sea and Pacific island (including parts of Micro, Polynesi) are not significantly different in the 2050 sea levels




